Modelling Stochastic Time Delay for Regression

### 2021-03-21 documentation/article/article.tex
Add references v1

\documentclass[11pt]{amsart}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[all]{xy}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{tabularx}
\bibliography{references}

@@

Consider the typical univariate regression problem where the intention is to find the relationship between $x$ and $y$. When extended to time series, a number of time specific complexities arise. For the specific case where the input $x$ affects the output $y$, with a random time delay in between, the estimated coefficients (or weights, predictions etc) are significantly attenuated. This attenuation occurs both in traditional statistical models and machine learning models, and for certain problems can be a serious limitation. The following document proposes techniques for handling this class of time series regression.

As an example, let us take the management of blood sugar level in diabetes patients. In people with diabetes, the pancreas can't effectively regulate blood sugar levels. Therefore, these levels must be controlled by insulin injections and a special diet. The challenge for many people is that the relationship between the input (insulin) and output (blood sugar) is extremely complex. The effect of the insulin injection might be observed after $15$, $20$  or $t$ minutes depending on a number of factors, many of which are unknown. Due to the stochastic nature of the time delays, the actual effect can't be easily determined. It is difficult to differentiate the effect of the insulin injection from other factors and accurately determine how much one should take.

@@

\section{Related Work}
There is extensive existing literature on time series problems with time delay dependencies. In the statistical disciplines there a number of treatments which focus on 
fixed time delay dependencies. For example, Granger Causality is used to determine whether one time series is useful for forecasting another, across some fixed delay.
Distributed Lag and Dynamic Regression models are able to handle linear and non-linear cause and effect relations that exist across multiple time lags.
Nonetheless, these models assume a fixed time lag dependency.

In the machine learning literature, there are a number of models which can handle complex dependencies across time. Sequence models such as the RNN, LSTM and GRU 
are able to generate predictions which incorporate time delays between input and the output variables. More recently, attention based models such as the Transformer have become the state-of-the-art for a
variety of tasks. Yet, all of these models have a tacit assumption of a fixed time lag dependency. There is also Dynamic Time Warping (DTW) which tries to find the optimal alignment between time series sequences by minimizing the
distance between the respective inputs. DTW is able to model varying time lag dependencies but is not purposed for regression, instead being mainly used for pattern matching and sequence alignment.

The field of system identification also has considerable literature on dynamic non-linear systems. For example, the NARMAX model can be used to identify 
non-linear systems with fixed time delays between input and output. At the same time, there are also treatments on systems with disturbances in the input, known as EIV systems. To our knowledge,
these works do not deal with uncertainty in the time domain.

Finally, recent papers such as Dynamic Time Lag Regression and Variable-Lag Granger Causality deal with non-stationary time lag dependencies. In other words, the lag structure is assumed to evolve over time.

@@

We consider the problem as analogous to the typical error-in-variables (EIV) regression. Ordinary regression analyses (and machine learning models) define the loss function with respect to errors in the $y$ axis only. For EIV, errors are considered in both the $y$-axis and the $x$-axis (REF). This is useful when there are measurement errors in the independent variable eg. because your physical measurements have some degree of random error. Similarly, for this problem we assume that we have errors in the $y$-axis and the $t$-axis. That is, there are random prediction errors and random errors in the time domain.

@@

\end{enumerate}
\vspace{5mm} %5mm vertical space
The procedure consists of an inner optimisation to find the set of $\mathcal{T}$ and an outer optimisation to find the set of parameters $\beta$, $\sigma_{\varepsilon}$ and $\sigma_{\tau}$. 
By iteratively alternating between the time shift optimisation and the parameter optimisation we are able to maximise equation (4). 
For the outer parameter optimisation, typical methods such as gradient descent, genetic algorithms and annealing can be used.  In our implementation, we have used the differential evolution algorithm (scipiy.optimize (REF)) because of its ability to handle noisy objective functions. In order to improve convergence, we also standardize all input variables to the range (0,1).

@@

In the univariate case, the utility of the described method is limited. However, we propose three extensions as future work. The first is to extend the method to multiple regression. We believe that the extension can be built on the same fundamental ideas presented in this document. Next, the model could be extended to include distributed lag structures. A distributed lag structure is where past values of the impulse influence future values of the output. Finally, in relating $x(t)$ to $f(x(t))$, the function $f$ could be extended to the modelling of non-linear forms.

@@

In addition, by incorporating these elements we could begin to explore alternative forms of time series prediction. Essentially, existing prediction mechanisms predict the average across time shifts.However, in some situations the predictions could be considered nonsensical.

@@

%\printbibliography

>>>>

\documentclass[11pt]{amsart}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[all]{xy}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{tabularx}

@@

Consider the typical univariate regression problem where the intention is to find the relationship between $x$ and $y$. When extended to time series, a number of time specific complexities arise. For the specific case where the input $x$ affects the output $y$, with a random time delay in between, the estimated coefficients (or weights, predictions etc) are significantly attenuated \cite{spearman}. This attenuation occurs both in traditional statistical models and machine learning models, and for certain problems can be a serious limitation. The following document proposes techniques for handling this class of time series regression.

As an example, let us take the management of blood sugar level in diabetes patients. In people with diabetes, the pancreas can't effectively regulate blood sugar levels. Therefore, these levels must be controlled by insulin injections and a special diet. The challenge for many people is that the relationship between the input (insulin) and output (blood sugar) is extremely complex \cite{blood_glucose}. The effect of the insulin injection might be observed after $15$, $20$  or $t$ minutes depending on a number of factors, many of which are unknown. Due to the stochastic nature of the time delays, the actual effect can't be easily determined. It is difficult to differentiate the effect of the insulin injection from other factors and accurately determine how much one should take.

@@

\section{Related Work}
There is extensive existing literature on time series problems with time delay dependencies. In the statistical disciplines there a number of treatments which focus on 
fixed time delay dependencies. For example, Granger Causality \cite{granger} is used to determine whether one time series is useful for forecasting another, across some fixed delay.
Distributed Lag \cite{almon} and Dynamic Regression models (e.g \cite{dynr}) are able to handle linear and non-linear cause and effect relations that occur across multiple time lags.
Nonetheless, these models assume a fixed time lag dependency.

In the machine learning literature, there are a number of models which can handle complex dependencies across time. Sequence models such as the RNN, LSTM and GRU \cite{ml}
are able to generate predictions which incorporate time delays between input and the output variables. More recently, attention based models such as the Transformer have become the state-of-the-art for a
variety of tasks, including time series forecasting \cite{deep_transformers}. Yet, all of these models have a tacit assumption of a fixed time lag dependency. There is also Dynamic Time Warping (DTW) \cite{dtw} which tries to find the optimal alignment between time series sequences by minimizing the
distance between the respective inputs. DTW is able to model varying time lag dependencies but is not purposed for regression, instead being mainly used for pattern matching and sequence alignment.

The field of system identification also has considerable literature on dynamic non-linear systems. For example, the NARX \cite{ml} model can be used to identify 
non-linear systems with fixed time delays between input and output. At the same time, there are also treatments on systems with disturbances in the input \cite{eiv_sysid}, known as EIV systems. To our knowledge,
these works do not deal with uncertainty in the time domain.

Finally, recent papers such as Dynamic Time Lag Regression \cite{dtlr} and Variable-Lag Granger Causality \cite{Amornbunchornvej_2019} deal with non-stationary time lag dependencies. In other words, the lag structure is assumed to evolve over time.

@@

We consider the problem as analogous to the typical error-in-variables (EIV) regression. Ordinary regression analyses (and machine learning models) define the loss function with respect to errors in the $y$ axis only. For EIV, errors are considered in both the $y$-axis and the $x$-axis \cite{deming}. This is useful when there are measurement errors in the independent variable e.g. because the physical measurements have some degree of random error. Similarly, for this problem we assume that we have errors in the $y$-axis and the $t$-axis. That is, there are random prediction errors and random errors in the time domain.

@@

\end{enumerate}
\vspace{5mm} %5mm vertical space
The procedure consists of an inner optimisation to find the set of $\mathcal{T}$ and an outer optimisation to find the set of parameters $\beta$, $\sigma_{\varepsilon}$ and $\sigma_{\tau}$. 
By iteratively alternating between the time shift optimisation and the parameter optimisation we are able to maximise equation (4). 
For the outer parameter optimisation, typical methods such as gradient descent, genetic algorithms and annealing can be used.  In our implementation, we have used the differential evolution algorithm \cite{diff_ev} (scipiy.optimize \cite{scipy}) because of its ability to handle noisy objective functions \cite{diff_ev}. In order to improve convergence, we also standardize all input variables to the range (0,1).

@@

In the univariate case, the utility of the described method is limited. However, we propose three extensions as future work. The first is to extend the method to multiple regression. We believe that the extension can be built on the same fundamental ideas presented in this document. Next, the model could be extended to include distributed lag structures. A distributed lag structure is where past values of the impulse influence future values of the output \cite{almon}. Finally, in relating $x(t)$ to $f(x(t))$, the function $f$ could be extended to the modelling of non-linear forms.

@@

In addition, by incorporating these elements we could begin to explore alternative forms of time series prediction. Essentially, existing prediction mechanisms predict the average across time shifts. However, in some situations the predictions could be considered nonsensical.

@@

\bibliography{references}
\bibliographystyle{ieeetr}
